Identification of protoneutron star g-modes in gravitational-wave data

### 2020-09-10 introduction.tex
Done with first two paragraphs of the intro

@@

The life of massive stars ($8 {\rm M}_\odot-100{\rm M}_\odot$) ends with the collapse of their iron core under their own gravity, leading the formation of a neutron star or a black hole (BH), followed (typically but not necessarily in the BH case) by the explosion of the star as a supernova. Core-collapse supernova (CCSN) explosions are one of the expected sources of gravitational-waves (GW) that have not yet been detected by current ground-based observatories. This is because even the most common type of CCSN, the neutrino-driven explosion, have a rate about three per century \cite{Gossan:2016} within our galaxy. The other main type of explosion, the magneto-rotational mechanism, can produce a more powerful signal and can be detected at distances up to $\sim 5$ Mpc \cite{Gossan:2016} \textcolor{red} Add better reference. However, the rate of events of this kind is much lower than the one for the neutrino driven mechanism $\sim 10^{-4} \rm{yr}^{-1}$, which represents less than $1 \%$ of all CCSNe.
Despite all this, collapsing stars produces a complex GW signal which could provide significant clues about the physical processes that occur in the moments after the collapse.

In the past years impressive progresses have been made in the development of numerical codes, which allow to obtain more accurate CCSN simulations. The waveforms produced by the magneto-rotational mechanism in particular is well understood. The core-bounce signal can be directly related with the rotational properties of the core \cite{Dimmelmeier:2007, abdikamalov:14, Richers:2017}. However, the low rate of this kind of events and its expected low amplitude in the slow-rotation case, will probably impede its detection.

In the case of the neutrino-driven explosion mechanism, the GW emission is mainly produced during the hydrodynamical bounce and the unstable evolution of the fluid inside the region formed by the recently formed proto-neutron star (PNS) and the accretion shock. The dynamics excite the different modes of oscillation of the PNS \cite{kokkotas, Friedman:2013}. Unluckily, in this case it is not posible to relate the GW emission with the properties (mass, rotation rate, metallicity or magnetic fields) of the progenitor stars.  A large number of physical processes are involved and their role is not completely understood. For instance,  uncertainties in the stellar evolution models of massive stars or in the nuclear and weak interactions necessary for the equation of state (EoS) of nuclear matter or the neutrino interactions. Furthermore, the stochastic and chaotic nature of the instabilities is transferred to the GW emission, resulting in the same progenitor leading significantly different waveforms.
The large number of physical ingredients in addition to the necessary accuracy of the modelling of complex multidimensional interactions requires large computational resources. One simulation of a single progenitor explosion  in 3D with accurate neutrino transport and realistic EoS can take several months of intense calculations on a scientific supercomputer facility. This complicates the systematic exploration of the progenitor parameters.

@@

In previous works \cite{Torres:2018, Torres:2019a}, we explore the eigenmode spectrum using results of CCSN numerical simulations and the theoretical model of the cavity form by the center of the PNS and the shock. We showed that the GW time-frequency distribution corresponds with the frequencies of oscillation of different families of p- and g-modes. These works reveal that is posible to perform CCSN asteroseismology and serves as a starting point to carry out inference of astrophysical parameters of PNSs. In this line of research, in \cite{Torres:2019b} we derived the relations between the different types of modes with some with the evolution of mass and radius of the PNS. These relations are universal in the sense that they not depend of the EOS, the mass of the progenitor or the code used to perform the simulation.

>>>>

% !TEX root = ccsn.tex

@@

The life of massive stars \pcd{(those born with masses between $\sim 8$~M$_\odot$ and $\sim 120$~M$_\odot$)} ends with the collapse of \pcd{the} iron core under \pcd{its} own gravity, leading \pcd{to} the formation of a neutron star \pcd{(NS)} or a black hole (BH), \pcd{and} followed (typically but not necessarily in the BH case) by \pcd{a supernova} explosion. \pcd{Nearby} core-collapse supernova (CCSN) explosions \pcd{are expected to be sources of gravitational waves (GW) and are one of the main candidates for the next great discovery by current ground-based observatories. However, these are relative rare events. A neutrino-driven explosion \citep{Bethe:1990} is the most likely outcome in the case of slow rotating cores, which are present in the bulk of CCSN progenitors. This event could be detected with advanced GW detectors within $5$~kpc \citep{Gossan:2016,TargetedSNSearchO12}. Such a galactic event has a rate of about $2-3$ per century \citep{Adams:2013,Rozwadowska:2021}. For the case of fast rotating progenitor cores the result is likely a magneto-rotational explosion, with } a more powerful signal \pcd{that could} be detected \pcd{within $50$~kpc and for some extreme models 
up to $5-30$ Mpc \cite{Gossan:2016,TargetedSNSearchO12}. However, only about $1\%$ of the electromagnetically observed events show signatures of fast rotation (broad-lined type Ic SNe \citep{Li:2011b} or events associated to long GRBs 
\citep{Chapman:2007}), making this possibility a subdominant channel of detection with an event rate of $\sim 10^{-4} \rm{yr}^{-1}$. Therefore, we focus this work only in neutrino-driven CCSNe. 
Despite the low rates, CCSN are of great scientific interest because they produce} a complex GW signals which could provide significant clues about the physical processes that occur in the moments after the collapse.

In the \pcd{last decade, a significant} progresses have been made in the development of numerical codes, \pcd{in particular in the treatment of multidimensioal effects \citep{BMueller:2020}.} \pcd{[PCD: I don't think this sentence is necessary since 
we have already discarded magneto-rotational events for this work. ]\sout{The waveforms produced by the magneto-rotational mechanism in particular is well understood. The core-bounce signal can be directly related with the rotational properties of the core} \cite{Dimmelmeier:2007, abdikamalov:14, Richers:2017}. \sout{However, the low rate of this kind of events and its expected low amplitude in the slow-rotation case, will probably impede its detection.}}

In the case of the neutrino-driven \pcd{explosions}, the GW emission is \pcd{primarly induced by instabilities developed at the newly formed proto-neutron star (PNS) and by the non-spherical accreting flow of hot matter over its surface.  This} dynamics excite the different modes of oscillation of the PNS\pcd{, which ultimately leads to the emission of GWs. [UNFINISHED! CONTINUE HERE]}Unluckily, in this case it is not posible to relate the GW emission with the properties (mass, rotation rate, metallicity or magnetic fields) of the progenitor stars.  A large number of physical processes are involved and their role is not completely understood. For instance,  uncertainties in the stellar evolution models of massive stars or in the nuclear and weak interactions necessary for the equation of state (EoS) of nuclear matter or the neutrino interactions. Furthermore, the stochastic and chaotic nature of the instabilities is transferred to the GW emission, resulting in the same progenitor leading significantly different waveforms.
The large number of physical ingredients in addition to the necessary accuracy of the modelling of complex multidimensional interactions requires large computational resources. One simulation of a single progenitor explosion  in 3D with accurate neutrino transport and realistic EoS can take several months of intense calculations on a scientific supercomputer facility. This complicates the systematic exploration of the progenitor parameters.

@@

\cite{kokkotas, Friedman:2013} In previous works \cite{Torres:2018, Torres:2019a}, we explore the eigenmode spectrum using results of CCSN numerical simulations and the theoretical model of the cavity form by the center of the PNS and the shock. We showed that the GW time-frequency distribution corresponds with the frequencies of oscillation of different families of p- and g-modes. These works reveal that is posible to perform CCSN asteroseismology and serves as a starting point to carry out inference of astrophysical parameters of PNSs. In this line of research, in \cite{Torres:2019b} we derived the relations between the different types of modes with some with the evolution of mass and radius of the PNS. These relations are universal in the sense that they not depend of the EOS, the mass of the progenitor or the code used to perform the simulation.
